# TODO Step 5: Share

In which I present all of my key findings and their supporting visualisations

> Most users don't get much exercise

Overall logged time:
```{r share_avg_time_by_intensity_viz}
print(viz_avg_time_by_intensity)
```
Active logged time:
```{r share_active_hrs_per_week_viz}
print(viz_active_hrs_per_week)
```
    
> For those who do exercise, intensity varied widely

* Proportion of Active time that is Very Active varies roughly evenly: viz_very_active_proportion
* More active users are only loosely related to higher-intensity users, with a wider range of times and percentages at the bottom than the top: viz_very_active_vs_total_active

> Users get around 7500 steps in per day on average, but this is varied

* viz_mean_daily_steps
    
> Walking is much more common than running:

* viz_mean_mod_walking, viz_mean_brisk_walking, viz_mean_running
  
> Users do not increase their step count, walking time, or walking intensity over time

* viz_mean_mod_walking_vs_time, viz_mean_brisk_walking_vs_time, viz_mean_running_vs_time
* viz_mean_daily_steps_vs_time

> Users prefer to engage in mobile exercise

* viz_distance_vs_activity (requires supporting explanation)

> Sleep logging was not common

* 

> Among those who did log, usage was spread out

* This is a similar trend to the exercise data

> Heart-rate data indicates overnight wear is split between low- and high-frequency users

> Device comfort may be a factor

> Consumer behaviour may be conducive to the Resting Heart Rate feature

* Existing consumers do not make much use of heart rate tracking overnight, despite this being a key part of the functionality of the Resting Heart Rate feature. Consumers do, however, log a lot of day-time data, and the majority of this data indicates low-intensity conditions
* Current user behaviors are therefore conducive to the Resting Heart Rate feature, but may result in higher-than-expected values as the RHR data will be skewed towards time spent awake rather than time spent asleep




## Exercise
### Key Findings

* Most people don't exercise much
* Those who do engage in a wide range of activity intensity levels
* Clear split between the majority, who occupy a wide range of the low end, and a few higher-intensity people
* Step counts average around 8k per day
* Mobile exercise in general appears common, but high-intensity running is not common
* Exercise quantity/quality did not improve over time

### Preliminary Findings

* The large majority of people's time is clearly being spent in a sedentary state.
* This makes sense, since it includes sleep tracking as well as time spent awake but immobile
* There is a wide spread of times spent non-sedentary, with no clear outliers at this point
* The large majority of non-sedentary time is Lightly Active

* 72.7% of the cohort (24 users) get less than 45 minutes of Active Time per week on average.
* 18% of the cohort (6 users) do more than 1 hour a week on average.
* Most people in the dataset are not doing much more than very light exercise.

* It's quite evenly spread out from around 10% Very Active to around 90% Very Active across all users, with no clear outliers.
* It would appear that the user group engages in a range of exercise activities of varying intensity.

* Very Active Proportion was positively correlated with Active time, with an R^2 value of 0.27.
* Removing the upper quintile decreased the correlation to 0.23, indicating that those doing less exercise also do a wider range of intensities
* Removing the lower quintile increased the correlation to 0.32, indicating that those doing the most exercise are doing a narrower range of higher-intensity activities.
* Overall, the results indicate that the majority of the cohort is engaged in a wide range of activities. There is a large subset of the group doing a low amount of varied exercise, as well as a small group of outliers doing a higher amount of more-intense exercise.

* Most people spend most of their time walking slowly
* Brisk walking is somewhat more common, with 24 of 33 users logging between 0.25 and 0.5 hours per week
* Runners are rare, with 30 of 33 users logging 0.25 hours or less per week, and the remaining three outliers logging between 0.5 and 1.25 hours per week

* Most users actually decreased or did not change their walking time over the data period
* This finding was consistent across all walking types
* There is no evidence to suggest FitBit users increased their walknig times as a result of their FitBit device usage

* As with walking types, the majority of users actually did fewer steps over time

* Active Minutes correlate strongly with Active distance, with R^2 = 0.77.
* Clearly absent from the plot is a significant number of users with high Active Minutes but low Active Distance
* Users typically log an additional 3.6 km of distance for every hour of Active time: to achieve this in a gym setting a user would have to follow a typical pattern resembling, for example, two one-minute stationary resistance training sets, followed by one minute moving around the gym at approximately 10.8 kilometers per hour, which seems unrealistic

* Overall, based on the strong correlation between Active Time and Active distance, and the scale of the distances covered by active users, it would appear that the cohort engages in a variety of mobile exercise activities. The exact breakdown of mobile vs. stationary activities has not been determined, but I believe it is safe to say that the marketing for the Ivy should represent mobile and outdoor activities more strongly over indoor, gym-based, or other stationary activities.

## Readiness Score
### Key Findings

* Sleep logging was not common
* Among those who did log, usage was spread out
* This is a similar trend to the exercise data
* Heart-rate data indicates overnight wear is split between low- and high-frequency users
* Device comfort may be a factor

### Preliminary Findings

* Sleep logging is available on all devices, so sleep log data should indicate how the overall cohort uses this feature
* Heart-rate logging is limited to later devices, so heartrate data will only indicate how that subset of the cohort uses that feature

* Out of all users in the data set, very few users are logging, with 36% (12 users) logging between 0% and 5% of nights, and a majority of 51% (17 users) logging 10% of nights or fewer
* Out of users with sleep log data, the results are more evenly spread, with 37% (9 users) logging 0% to 15% of nights, and another 42% (10 users) logging 70% to 90% of nights.

* Out of all users, 73% (24 users) did not log any heartrate-based sleep logs.
* Out of all users, 42% (14 users) logged heart-rate data
* Out of those users, the results were split, with 50% (7 users) logging sleep on 35% to 90% of nights, and the other 50% logging sleep on 0% to 15% of nights
* The heart-rate-based analysis produces the same results for individual users as the sleep-log-based analysis
* Overall, the results reinforce the findings of the sleep-log analysis, but only for the smaller sub-set of users that have devices with heart-rate tracking.

## Sleep Tracking
### Key Findings

* Sleep quantity/quality started high
* Typically did not improve over time

### Preliminary Findings

* The results appear realistic: the mean and median are 6.9 and 7.4 hours, respectively, with 79% of users (19 users) getting between 6 and 9 hours of sleep on average

* The majority of users appear to be getting high-quality sleep, with 88% of users logging between 90% and 95% of time as Asleep
* Time spent Awake is almost nil, with 88% of users logging less than 5% of time asleep. This is likely due to the programming of the FitBit devices, which feature an automatic sleep logging function that kicks in when it detects the user may be asleep, based on movement and heart rate data

* The correlations for sleep quantity and quality formed approximately-normal distributions around zero, indicating no consistent increase for either variable
* Sleep quantity appeared to decrease slightly, with mean and median correlations of -0.09 and -0.13, respectively
* Sleep quantity was approximately static, with mean and median correlations of -0.04 and -0.02, respectively
* Overall, there is nothing to indicate that users increased the quantity or quality of their sleep over time
* The quantity and quality of sleep for this cohort of users was already relatively good, with the majority of users getting between 6-9 hours sleep, and spending upwards of 90% of that time in a stable Asleep state.



## Resting Heart Rate and Cardiac Coherence
### Key Findings

* Existing consumers do not make much use of heart rate tracking overnight, despite this being a key part of the functionality of the Resting Heart Rate feature
* Consumers do, however, log a lot of day-time data, and the majority of this data indicates low-intensity conditions
* Current user behaviors are therefore conducive to the Resting Heart Rate feature, but may result in higher-than-expected values as the RHR data will be skewed towards time spent awake rather than time spent asleep

### Preliminary Findings

* Daytime logging is much more popular than nighttime logging
* 50% of users (7 users) logged 6 hours on between 80% to 95% of days
* The other 50% of users logged on between 0% and 75% of days

* The majority of logs for all but one user are in the "Sedentary" category, which equates to less than 95 bpm
* It is fair to assume therefore that users are not only using their heart-rate trackers for exercise
* There should be ample data to determine a user's resting heart rate from the daytime data, even if there is little nighttime data

## Features Without Data
### Key Findings

Several functions and features of the Ivy are not present on the FitBits in the data set. Given the lack of data, it's not possible to directly analyse existing user behaviours with respect to these features, which include:

* Respiratory Rate
* Hydration tracking
* Mindfulness Minute Tracking
* Menstrual Cycle
* Wellness Score

# TODO: Step 6: Act
## Exercise Recommendations

* To appeal to the FitBit market, the marketing for the Ivy's exercise tracking should highlight a broad range of activities ranging from gentle to more intense. The marketing should not target highly-athletic groups, as these represent only a small part of the user base.
* The marketing for the Ivy should avoid any claims that users will increase their activity levels or their step counts as a result of their purchasing the device, as the data does not support this claim.
* The cohort does appear to engage in outdoor activities that cover long distances such as running, as opposed to stationary activities such as gym workouts, so the marketing can highlight activities like light jogging, walking dogs, and cycling

## Readiness Score Recommendations

* Overnight usage of FitBits appears to be split between a large majority who rarely wear their devices overnight, and a smaller minority who wear more frequently, up to 90% of the time
* The Readiness Score can therefore be marketed as a value-add feature that produces more useful information when worn to bed than competing products
* The non-allergenic materials used in the Ivy should be used to highlight the comfort aspects of the device when marketing features that require overnight wearing, as discomfort from overnight wear may be one of the factors contributing to the low utilisation of sleep logging in existing products.

## Sleep Tracking Recommendations

* Don't market this feature as something that will improve your sleep
* Instead, given the consistency of users in getting their sleep, as evidenced by the small correlation with time for sleep quantity and quality, the Sleep Tracking could be marketed more as a way of tracking and reinforcing the user's existing good sleep habits 
* As with the Readiness Score, marketing teams could frame the feature more as a way of getting a heads-up on days where you haven't slept well, since most of the time the feature would just tell users (from this cohort, at least) that they just had a reasonably good night's sleep. This would work well with the "track existing habits" marketing angle: think "You're great at giving your body the sleep it needs, but for those nights where things just don't work out, Ivy can give you guidance on whether you're up for the day, or better off taking it down a notch to recover."
* If it exists, highlight a manual-activation mode for the Sleep Tracking mode. The low amount of non-Asleep data in this data set may be a result of people mostly using the automatic mode, which is convenient, but doesn't track time spent *trying* to get to sleep: a manual-activation feature could therefore be pushed harder as a sort of second feature that gets better data on how you go getting to sleep.

## Resting Heart Rate and Cardiac Coherence Recommendations

* In order to match current user behaviours, the marketing for the Resting Heart Rate should emphasise daytime usage: for instance, advertise that the feature will help understand your resting heart rate during your work day, 
* The data indicates that users typically do not log overnight, which may impact the performance of this feature: part of the usefulness of the feature comes from calibrating the data based on overnight heart-rate, so if users are not wearing overnight, the resting heart rate value may be skewed upward s as it is based primarily on daytime usage
* The marketing should therefore encourage users to wear overnight, highlighting the synergy between this and the Readiness Score feature to maximise the likelihood that users wear overnight, and therefore the accuracy of both features
* If users do not wear their devices overnight, they are unlikely to see the benefits and may consider other, less-featured but cheaper devices for their next purchase

## Features without equivalent data Recommendations

* These features can all be marketed as improvements and additions over the functionality of existing FitBits.
* As discussed with regards to the Readiness score, marketing for any feature that requires overnight wear, like the Cardiac Coherence tracking, should take into account the low utilisation of existing overnight functions like sleep tracking seen in the existing data.

